---
title: "Applications - Single DL"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(multipleDL)
library(tidyverse)
library(kableExtra)
library(quantreg)
library(survival)
```

```{r}
load('synthetic_data_single.Rda')
```

## Data

- `r nrow(data)` subjects

## Outcome 

- IL-4 
    
    - Range `r summary(data$il_4)[1] %>% round(3)` to `r summary(data$il_4)[6] %>% round(3)`
    
    - Lower DL at 0.047
    
    - `r is.na(data$il_4) %>% sum` below DL (`r (((is.na(data$il_4) %>% mean) * 100)) %>% round()`%)

## Covariates

- BMI

- Age

- Sex 

  - Female
  
  - Male

- HIV status

  - Group 1: HIV+ insulin sensitive (HbA1c < 5.7% or FBG < 100)
  
  - Group 2: HIV+ pre-diabetic (HbA1c 5.7-6.4% or FBG 100-126)
  
  - Group 3: HIV+ diabetic
  
  - Group 4: HIV-neg diabetic
  
Distribution of outcomes [Figure S6 (A) (B)]
```{r}
par(mfrow=c(1,2))
hist(data %>% filter(dl==1) %>% select(il_4) %>% pull, breaks=30, xlab = 'IL-4', main = 'Historgram of IL-4')
hist(data %>% filter(dl==1) %>% select(il_4) %>% pull %>% log, breaks=30, xlab = 'IL-4', main = 'Historgram of log(IL-4)')
```


# Model

## CPM 

### Odds ratio [Table S9]
```{r}
mod <- multipleDL(il_4 ~ age+ sex + bmi + status,
                  data = data, delta_lower = data$dl, link='logit')

beta <- mod$coef[!grepl("^alpha", names(mod$coef))]
beta.se <- (mod$var %>% diag %>% sqrt)[!grepl("^alpha", names(mod$coef))]

# odds ratio
or <- exp(beta)
# odds ratio lower bound
or_lb <- exp(beta - qnorm(0.975) * beta.se)
# odds ratio upper bound
or_ub <- exp(beta + qnorm(0.975) * beta.se)
# p-value
pval <- (1 - pnorm(abs(beta / beta.se))) * 2


# table for odds rato
tab_or <- data.frame(or = or, or_lb = or_lb, or_ub = or_ub, pval)

# age (per 10 years)
tab_or[1, 'or'] <- exp(beta[1] * 10)
tab_or[1, 'or_lb'] <- exp((beta[1] - qnorm(0.975) * beta.se[1]) * 10)
tab_or[1, 'or_ub'] <- exp((beta[1] + qnorm(0.975) * beta.se[1]) * 10)

# BMI (5 unit)
tab_or[3, 'or'] <- exp(beta[3] * 5)
tab_or[3, 'or_lb'] <- exp((beta[3] - qnorm(0.975) * beta.se[3]) * 5)
tab_or[3, 'or_ub'] <- exp((beta[3] + qnorm(0.975) * beta.se[3]) * 5)

tab_or %>% 
  kable(digits=3) %>% 
  kable_styling(full_width = F)
```
### Transformation [Figure S6 (C)]
```{r}
plot(mod$yunique, c(mod$coef[1:length(mod$yunique)-1], Inf),
     type = 'l', xlab = 'IL-4 (pg/ml)',  ylab=expression(paste(hat(alpha), "(y)")),
     xlim = c(0.045, 0.06))
```

### Conditional quantities [Figure S7(A), Figure S8(A)(B)(C)]
```{r}
new <- 31:41 # bmi
new.data <- matrix(NA, ncol = mod$p, nrow=length(new))
colnames(new.data) <- names(beta)

new.data[,1] <- median(data %>%  pull(age))
new.data[,2] <- 1 # male
new.data[,3] <- new
new.data[,4] <- 0 # group 1
new.data[,5] <- 0 # group 1
new.data[,6] <- 0 # group 1

# quantiles
q50.est <- quantile_dl(mod, new.data, 0.5)
q90.est <- quantile_dl(mod, new.data, 0.9)

# cdf
c003.est <- cdf_dl(mod, new.data, 0.047)
c005.est <- cdf_dl(mod, new.data, 0.055)

par(mfrow=c(2,2))
plot(new, q50.est[[1]],
     cex= 0.01, ylim = c(0.045, 0.06),
     ylab="IL-4", xlab="BMI", type="n", main="50th percentile", yaxt='n')
axis(2, las=2, at=c(0.047, c(0.05, 0.06)), label=c("<0.047", c(0.05, 0.06)))
polygon(c(new, rev(new)), c(q50.est[[2]], rev(q50.est[[3]])), col = "grey", border = NA)
lines(new, q50.est[[1]], cex=0.01, lty=1)
lines(new, q50.est[[2]], cex=0.01, lty=2)
lines(new, q50.est[[3]], cex=0.01, lty=2)

plot(new, q90.est[[1]],
     cex= 0.01, ylim = c(0.045, 0.065),
     ylab="IL-4", xlab="BMI", type="n", main="90th percentile", yaxt='n')
axis(2, las=2, at=c(0.047, c(0.05, 0.06)), label=c("<0.047", c(0.05, 0.06)))
polygon(c(new, rev(new)), c(q90.est[[2]], rev(q90.est[[3]])), col = "grey", border = NA)
lines(new, q90.est[[1]], cex=0.01, lty=1)
lines(new, q90.est[[2]], cex=0.01, lty=2)
lines(new, q90.est[[3]], cex=0.01, lty=2)


plot(new, 1-c003.est[[1]],
     cex= 0.01, ylim=c(0.4, 1),
     ylab="Probability", xlab="BMI", type="n", main="P(IL-4 > 0.047 | X)", yaxt='n')
axis(2, las=2, at=seq(0.4, 1, 0.2), label=seq(0.4, 1, 0.2))
polygon(c(new, rev(new)), c(1-c003.est[[3]], rev(1-c003.est[[4]])), col = "grey", border = NA)
lines(new, 1-c003.est[[1]], cex=0.01)
lines(new, 1-c003.est[[3]], cex=0.01, lty=2)
lines(new, 1-c003.est[[4]], cex=0.01, lty=2)


plot(new, 1-c005.est[[1]],
     cex= 0.01, ylim=c(0, 0.62),
     ylab="Probability", xlab="BMI", type="n", main="P(IL-4 > 0.055 | X)", yaxt='n')
axis(2, las=2, at=seq(0, 0.6, 0.2), label=seq(0, 0.6, 0.2))
polygon(c(new, rev(new)), c(1-c005.est[[3]], rev(1-c005.est[[4]])), col = "grey", border = NA)
lines(new, 1-c005.est[[1]], cex=0.01)
lines(new, 1-c005.est[[3]], cex=0.01, lty=2)
lines(new, 1-c005.est[[4]], cex=0.01, lty=2)
```


## Median regression [Figure S7(C)]
```{r}
# median
mod_rq <- rq(il_4  ~ age + sex + bmi + status,
             data = data)

X <-  as.matrix(cbind(1, new.data))
pred <-drop(X %*% mod_rq$coef)
V <- summary(mod_rq, cov = TRUE)
sdpred <- sqrt(diag(X %*% V$cov %*% t(X)))
med_est <- pred
med_lower <- pred - qnorm(0.975) * sdpred
med_upper <- pred + qnorm(0.975) * sdpred

plot(new, med_est,
     cex= 0.01, ylim =  c(0.05, 0.06),
     ylab="IL-4", xlab="BMI", type="n", main="50th percentile", yaxt='n')
axis(2, las=2, at=seq(0.05, 0.06, 0.005), label=seq(0.05, 0.06, 0.005))
polygon(c(new, rev(new)), c(med_lower, rev(med_upper)), col = "grey", border = NA)
lines(new, med_est, cex=0.01, lty=1)
lines(new, med_lower, cex=0.01, lty=2)
lines(new, med_upper, cex=0.01, lty=2)
```

## Likelihood Approach [Figure S7(B)]
```{r}
mod_mle <- survreg(Surv(data$il_4, data$dl, type='left') ~ age + sex + bmi + status, 
                   data = data, dist = 'lognormal')

new.data <- data.frame(age = median(data %>% pull(age)),
                       sex = 'Male',
                       bmi = new,
                       status = "Group 1: HIV+ insulin sensitive (HbA1c < 5.7% or FBG < 100)")

# median (censored quantile regression)
med_mle <- predict(mod_mle, newdata=new.data, 
                   type = "quantile", se=T, p=0.5)

med_est2 <- med_mle$fit
med_lower2 <- med_mle$fit - qnorm(0.975) * med_mle$se.fit
med_upper2 <- med_mle$fit + qnorm(0.975) * med_mle$se.fit

plot(new, med_est2,
     cex= 0.01, ylim =  c(0.04, 0.06),
     ylab="IL-4", xlab="BMI", type="n", main="50th percentile", yaxt='n')
axis(2, las=2, at=seq(0.04, 0.06, 0.01), label=seq(0.04, 0.06, 0.01))
polygon(c(new, rev(new)), c(med_lower2, rev(med_upper2)), col = "grey", border = NA)
lines(new, med_est2, cex=0.01, lty=1)
lines(new, med_lower2, cex=0.01, lty=2)
lines(new, med_upper2, cex=0.01, lty=2)
```
